Efficient Calculation of Polynomial and Interaction Features on Sparse Matrices

### 2015-10-02 paper/paper2.tex
fixed row index in algo
properly slicing into nz cols
two cols and all figs eps

\documentclass[11pt,onecolumn]{article}

@@

%Same as above, but in vector graphic form. Use this in the final.
%\begin{sidewaysfigure}
%    \centering
%    \includegraphics[scale=0.65]{classifier_boundaries.pdf}
%    \caption{Decision boundaries for four classifier types on four different datasets. The darker the color, the more certain the decision.
%             The plain dataset is shown in the first column. Each other column is a type of classifier applied to that dataset. The classifier
%             types are logistic regression, logistic regression with second degree polynomial features, RBF SVM, and gradient boosting with decision trees.
%             The accuracy of each model is shown in the bottom right corner of that model's subplot.}
%    \label{fig:boundaries}
%\end{sidewaysfigure}

\begin{sidewaysfigure*}[ht!, numbered]
    \centering
    \includegraphics[scale=0.5]{classifier_boundaries.png}
    \caption{Decision boundaries for four classifier types on four different datasets. The darker the color, the more certain the decision.
             The plain dataset is shown in the first column. Each other column is a type of classifier applied to that dataset. The classifier
             types are logistic regression, logistic regression with second degree polynomial features, RBF SVM, and gradient boosting with decision trees.
             The accuracy of each model is shown in the bottom right corner of that model's subplot.}
    \label{fig:boundaries}
\end{sidewaysfigure*}

@@

\begin{codebox}
\Procname{$\proc{Dense Polynomial Features}(A)$}
    \zi $N \gets$ row count of $A$
    \zi $D \gets$ column count of $A$
    \zi $B$ $\gets$ Matrix of size $N \times \frac{D^2+D}{2}$
    \zi \For $\id{row}$ in $A$ \Do
    \zi     $k \gets 0$
    \zi     \For $\i \gets 0 \To D$ \Do
    \zi         \For $\j \gets i \To D$ \Do
    \zi             $B[i, k] \gets \id{row}[i] \cdot \id{row}[j]$
    \zi             $k \gets k + 1$
                \End
            \End
       	\End
\end{codebox}

@@

\begin{codebox}
\Procname{$\proc{Sparse Polynomial Features}(A)$}
    \zi $\func{PolyMap}(a, b) = \frac{2Da-a^2+2b-3a-2}{2}+a+1$
    \zi $N \gets$ row count of $A$
    \zi $D \gets$ column count of $A$
    \zi $B$ $\gets$ Compressed Sparse Row Matrix of size $N \times \frac{D^2+D}{2}$
    \zi \For $\id{row}$ in $A$ \Do
    \zi     $N_{zc} \gets$ nonzero columns of $row$
    \zi     \For $i \gets 0 \To |N_{zc}|$ \Do
    \zi         \For $j \gets i \To |N_{zc}|$ \Do
    \zi             $k \gets \func{PolyMap}(i, j)$
    \zi             $r \gets$ index of $\id{row}$
    \zi             $B[r, k] \gets \id{row}[i] \cdot \id{row}[j]$
                \End
            \End
       	\End
\end{codebox}

@@

\begin{codebox}
\Procname{$\proc{Sparse Interaction Features}(A)$}
    \zi $\func{InterMap}(a, b) = \frac{2Da-a^2+2b-3a-2}{2}$
    \zi $N \gets$ row count of $A$
    \zi $D \gets$ column count of $A$
    \zi $B$ $\gets$ Compressed Sparse Row Matrix of size $N \times \frac{D^2-D}{2}$
    \zi \For $\id{row}$ in $A$ \Do
    \zi     $N_{zc} \gets$ nonzero columns of $row$
    \zi     \For $i \gets 0 \To |N_{zc}|-1$ \Do
    \zi         \For $j \gets i+1 \To |N_{zc}|$ \Do
    \zi             $k \gets \func{InterMap}(i, j)$
    \zi             $r \gets$ index of $\id{row}$
    \zi             $B[r, k] \gets \id{row}[i] \cdot \id{row}[j]$
                \End
            \End
       	\End
\end{codebox}

@@

\subsection{Empirical}
To assess the performance of the algorithm on data of varied density, we compare its speed
to the PolynomialFeatures class of the popular machine learning scikit-learn \cite{scikit-learn}.
Uniformly random matrices of size $1000 \times D$ were generated for $d=0.2$, $0.4$, $0.6$, and $0.8$, and 
$D=1$, $101$, $201$, $301$, $401$, and $501$. Since the scikit-learn version of the algorithm
does not exploit sparsity, we assume the method's time is invariate with respect the density of the matrix, but
we gave it the lowest density of each dimensionality to be fair. The results are shown in figure \eqref{fig:benchmark} on page \pageref{fig:benchmark}.

\begin{figure*}[ht!, numbered]
    \includegraphics[scale=0.3]{benchmark_results.png}
    \caption{This plot shows how long the sparse polynomial method took to run on matrices with 1000 rows of various dimensionalities and densities and compared to the scikit-learn PolynomialFeatures class on various dimensionalities.}
    \label{fig:benchmark}
    \centering
\end{figure*}

@@

\begin{table}
    \centering
    \begin{tabular}[numbered]{c | c | c} \label{table:results}
 %       \hline
        dataset             & 20newsgroups  & connect-4 \\ \hline
        instances           & 11314        & 67557    \\ \hline
        features            & 130107       & 126       \\ \hline
        polynomial features & 8463980778 & 8,127     \\ \hline
        density             & 0.12          & 0.11      \\ \hline
        dense space (MB)    & Memory Error  & 4191      \\ \hline
        sparse space (MB)   & 5333          & 735       \\ \hline
        dense time (s)      & Memory Error  & 26        \\ \hline
        sparse time (s)     & 109           & 44        \\ \hline
%        \hline
    \end{tabular}
    \caption{Time and space comparison of second degree polynomial feature calculation between the sparse method and scikit-learn.}
\end{table}

>>>>

\documentclass[11pt,twocolumn]{article}

@@

To demonstrate the benefit of polynomial features in regression, we two ordinary least squares models to $y=sin(x) + \mathcal{N}(0, 0.1), x \in [0, \pi]$. The first model (the blue curve) was given only $x$ as its input. The second model (the green curve) was given $x$ and second degree polynomial features derived from $x$ (namely $x^2$). The curves the models learned can be seen in figure \ref{fig:regression}.

\begin{sidewaysfigure}
    \centering
    \includegraphics[scale=0.3]{boundaries.eps}

     \caption{\footnotesize{Decision boundaries for four classifier types on four different datasets. The darker the color, the more certain the decision.
             The plain dataset is shown in the first column. Each other column is a type of classifier applied to that dataset. The classifier
             types are logistic regression, logistic regression with second degree polynomial features, RBF SVM, and gradient boosting with decision trees.
             The accuracy of each model is shown in the bottom right corner of that model's subplot.}}
    \label{fig:boundaries}
\end{sidewaysfigure}

\begin{figure}
    \centering
    \includegraphics[scale=0.4]{regression.eps}
     \caption{\footnotesize{Two linear regression models fit to $y=sin(x) + \mathcal{N}(0, 0.1), x \in [0, \pi]$. The blue line's model was given only $x$, while the green line's model was given $x$ and second degree polynomial features.}}
    \label{fig:regression}
\end{figure}

%\newpage

@@

\begin{codebox}
\footnotesize
\Procname{$\proc{Dense Polynomial Features}(A)$}
    \zi $N \gets$ row count of $A$
    \zi $D \gets$ column count of $A$
    \zi $B$ $\gets$ Matrix of size $N \times \frac{D^2+D}{2}$
    \zi \For $\id{row}$ in $A$ \Do
    \zi     $k \gets 0$
    \zi     \For $\i \gets 0 \To D$ \Do
    \zi         \For $\j \gets i \To D$ \Do
    \zi             $r \gets$ index of $\id{row}$
    \zi             $B[r, k] \gets \id{row}[i] \cdot \id{row}[j]$
    \zi             $k \gets k + 1$
                \End
            \End
       	\End
\end{codebox}

@@

\begin{codebox}
\footnotesize
\Procname{$\proc{Sparse Polynomial Features}(A)$}
    \zi $\func{PolyMap}(a, b) = \frac{2Da-a^2+2b-3a-2}{2}+a+1$
    \zi $N \gets$ row count of $A$
    \zi $D \gets$ column count of $A$
    \zi $B$ $\gets$ empty sparse $N \times \frac{D^2+D}{2}$ matrix
    \zi \For $\id{row}$ in $A$ \Do
    \zi     $N_{zc} \gets$ nonzero columns of $row$
    \zi     \For $i \gets 0 \To |N_{zc}|$ \Do
    \zi         \For $j \gets i \To |N_{zc}|$ \Do
    \zi             $k \gets \func{PolyMap}(N_{zc}[i], N_{zc}[j])$
    \zi             $r \gets$ index of $\id{row}$
    \zi             $B[r, k] \gets \id{row}[N_{zc}[i]] \cdot \id{row}[N_{zc}[j]]$
                \End
            \End
       	\End
\end{codebox}

@@

\begin{codebox}
\footnotesize
\Procname{$\proc{Sparse Interaction Features}(A)$}
    \zi $\func{InterMap}(a, b) = \frac{2Da-a^2+2b-3a-2}{2}$
    \zi $N \gets$ row count of $A$
    \zi $D \gets$ column count of $A$
    \zi $B$ $\gets$ empty sparse $N \times \frac{D^2-D}{2}$ matrix
    \zi \For $\id{row}$ in $A$ \Do
    \zi     $N_{zc} \gets$ nonzero columns of $row$
    \zi     \For $i \gets 0 \To |N_{zc}|-1$ \Do
    \zi         \For $j \gets i+1 \To |N_{zc}|$ \Do
    \zi             $k \gets \func{InterMap}(N_{zc}[i], N_{zc}[j])$
    \zi             $r \gets$ index of $\id{row}$
    \zi             $B[r, k] \gets \id{row}[N_{zc}[i]] \cdot \id{row}[N_{zc}[j]]$
                \End
            \End
       	\End
\end{codebox}

@@

\subsection{Empirical}
To assess the performance of the algorithm on data of varied density, we compare its speed
to the PolynomialFeatures class of the popular machine learning scikit-learn \cite{scikit-learn}.
Uniformly random matrices of size $100 \times D$ were generated for $d=0.1$, $0.2$, $0.3$, $0.4$, and $0.5$, and 
$D=1$, $101$, $201$, $301$, $401$, and $501$. Each point was resampled five times and averaged to decrease variance. Since the scikit-learn version of the algorithm
does not exploit sparsity, we assume the method's time is invariate with respect the density of the matrix, but
we gave it the lowest density of each dimensionality to be fair. The results are shown in figure \eqref{fig:benchmark} on page \pageref{fig:benchmark}.

\begin{figure}[ht, numbered]
    \includegraphics[scale=0.4]{benchmark.eps}
    \caption{This plot shows how long the sparse polynomial method took to run on matrices with 1000 rows of various dimensionalities and densities and compared to the scikit-learn PolynomialFeatures class on various dimensionalities.}
    \label{fig:benchmark}
    \centering
\end{figure}

@@

\begin{table}
    \centering
    \small    
    \begin{tabular}[numbered]{c | c | c} \label{table:results}
 %       \hline
        dataset             & 20newsgroups  & connect-4 \\ \hline
        instances           & 11,314        & 67,557    \\ \hline
        features            & 130107       & 126       \\ \hline
        polynomial features & 8,463,980,778 & 8,127     \\ \hline
        density             & 0.12          & 0.11      \\ \hline
        dense space (MB)    & Memory Error  & 4191      \\ \hline
        sparse space (MB)   & 5333          & 735       \\ \hline
        dense time (s)      & Memory Error  & 26        \\ \hline
        sparse time (s)     & 109           & 44        \\ \hline
%        \hline
    \end{tabular}
    \caption{Time and space comparison of second degree polynomial feature calculation between the sparse method and scikit-learn.}
\end{table}
